Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-23 paper.qmd
more intro to nix

The [Nix]{.pkg} package manager provides these capabilities. [Nix]{.pkg} ensures
reproducible software installation by deploying *component closures*—packages
bundled with all their dependencies and transitive dependencies
[@dolstra2004nix]. This creates self-contained software environments that work
identically across machines and over time. [Nix]{.pkg} can replace
[Docker]{.pkg} for environment isolation, [renv]{.pkg} for package management,
and even [Make]{.pkg} for workflow orchestration—all within a single, unified
framework.

@@

Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports
reproducibility. [Nix]{.pkg} is a *functional* package manager in the
programming sense: it treats software builds as pure functions. Given the same
inputs (source code, dependencies, build scripts), a [Nix]{.pkg} build will
always produce the identical output, regardless of when or where it's executed.
This is achieved by disallowing side effects and ensuring that builds operate
within hermetic (isolated) environments, with no hidden dependencies on global
system state. While this functional approach greatly enhances reproducibility,
it can sometimes introduce complexity for package maintainers, especially for
software that needs to download external assets during installation (e.g.,
certain Bioconductor packages). For end-users, however, this complexity is
largely abstracted away. Additionally, [Nix]{.pkg} supports multiple versions or
*variants* of a package on the same system. Each package installed into the "Nix
store" (typically in `/nix/store`) has a unique identifier derived from a hash
of all its inputs. This isolation ensures that projects always use their
intended software versions, allowing multiple versions of [R]{.proglang} to
coexist and preventing "dependency hell" and global environment pollution.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`. This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.

>>>>

The [Nix]{.pkg} package manager provides these capabilities. [Nix]{.pkg} ensures
reproducible software installation by deploying *component closures*—packages
bundled with all their dependencies and transitive dependencies
[@dolstra2004nix]. Think of installing a package like packing for a trip.
Traditional package managers are like packing just your clothes, assuming you'll
find a toothbrush and soap at your destination. [Nix]{.pkg} packs everything you
might need—the entire self-contained kit. This is what [Nix]{.pkg} calls a
*component closure*.

This creates self-contained software environments that work
identically across machines and over time. [Nix]{.pkg} can replace
[Docker]{.pkg} for environment isolation, [renv]{.pkg} for package management,
and even [Make]{.pkg} for workflow orchestration—all within a single, unified
framework.

@@

Pinning is essential, but it is not the only mechanism through which [Nix]{.pkg}
achieves reproducibility. [Nix]{.pkg} adopts principles from functional
programming, conceptualizing software builds as *pure functions*. In
mathematics, a pure function is defined as one that always produces the same
output for a given input—for instance, if \( f(x) = x^2 \), then \( f(2) = 4 \)
invariably holds. Analogously, [Nix]{.pkg} ensures that, given identical inputs
(such as source code, dependencies, and build instructions), the resulting
output (the built package) remains identical, irrespective of the environment or
time of execution. In this sense, [Nix]{.pkg} functions as a *functional package
manager*, applying the deterministic properties of pure functions to software
construction.

This is achieved by disallowing side effects and ensuring that builds operate
within hermetic (isolated) environments, with no hidden dependencies on global
system state. While this functional approach greatly enhances reproducibility,
it can sometimes introduce complexity for package maintainers, especially for
software that needs to download external assets during installation (e.g.,
certain Bioconductor packages). For end-users, however, this complexity is
largely abstracted away. Additionally, [Nix]{.pkg} supports multiple versions or
*variants* of a package on the same system.

[Nix]{.pkg} installs all packages into a centralized, content-addressed
directory called the [Nix]{.pkg} store (typically `/nix/store`). Each package
gets a unique subdirectory whose name includes a hash of all its inputs: source
code, dependencies, build instructions. This means
`/nix/store/abc123-dplyr-1.0.0` and `/nix/store/xyz789-dplyr-1.1.0` can coexist
without conflicts.

This isolation ensures that projects always use their intended software
versions, allowing multiple versions of [R]{.proglang} to coexist and preventing
"dependency hell" and global environment pollution.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, I recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`, which executes following steps:

- Nix reads the expression and identifies what needs to be installed;
- It checks if these packages already exist in `/nix/store`;
- If not, it downloads or builds them (with all dependencies);
- It creates a new shell environment and updates the `$PATH` to include the
  right binaries from `/nix/store`.
- The user is now in an interactive [Nix]{.pkg} shell session.

This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.
